---
title: "DuckDB and R for Data Analysis"
author: "Dr Pietà-Georgina 'Georgie' Schofield"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: false
    highlight: zenburn
  github_document:
    toc: true
---

```{r setup_R, comment=NA, echo=F}
require(tidyverse)
require(dbplyr)
.locData <- file.path(Sys.getenv("HOME"),"Projects","demrfmr",".data")
dbname <- file.path(.locData,"duckdb","demrisk.duckdb")
tohtml <- FALSE
```

# Motivation

## CPRD Aurum Data

- 250000 dementia patient records from CPRD Aurum
- 13GB of data in 7 compressed files (each ~2GB)
- Unzips to 82GB of data
    - patient: 249961 records in 1 file
    - practice: 1483 records in 1 file
    - staff: 1336287 records in 1 file
    - consultation: 120456179 records in 10 files
    - observation: 425006899 records in 43 file
    - referral: 2826888 records in 1 file
    - problem: 14501709 records in 1 file(s)
    - drugissue: 197222175 records in 35 

# Problem

## Find Index Age Distribution by Dementia Type

- You have a laptop and R and limited memory
- You need to find index date for a number of codes groups representing different dementia types
- R freezes or just crashes if you try to load 425 Million Records from 43 files
- You could load each in turn and filter save but...
    - What about patients spanning more than one file
    - At some point you need to realise there are data quality issues to sort out
    - Repeat the whole process

# A Solution 

## `duckdb`, `tidyverse`, `DBI` and `dbplyr`

- `tidyverse`: (`tibble`, `dplyr`, `ggplot2`, `lubridate` etc.) 
    - Other alternative object paradigms exist data.tables but I am at home in the `tidyverse` 
- duckdb: is a column orientated local file based `sql` database engine
    - specially tuned for analytics (OLAP) over transaction processing (OLTP)
    - similar but much better data science alternative to SQLite
- `DBI`: connect R to databases 
    - `dbplyr` permits use of `tidyverse` functions of database tables rather than `tibbles`
    - can use SQL queries that is your preference.

# Example ETL

## Extract

- Well I have done that and you have your 9 zipped files (13GB)
- Copy them from CPRDGeneralinfo drive to a University asset you have read/write access to
    - options university desktop / laptop / active data storage network drive (preferable)
- Unzip the compressed files into the 93 tab delimited text files (82GB)

## Load

- Download my `rtrhd` R package (GitHub) (NB. It has a lot of irrelevant functions)
- get a copy of the CPRD Aurum data-dictionary file from CPRDGeneralinfo drive
- Then run this 

<div style="font-size: 0.8em;">
```{r load_R, comment=NA, eval=F}
# You will need the location of your files here so I have just put variables
rtrhd::load_with_data_dictionary(
  ddFile=ddfile, # Variable holding location of data-
                 # dictionary file
  dbf=dbname, # Variable holding location of duckdb database file 
              # to be created
  datadir=ddir, # Variable pointing to parent directory where the
                # text files were unzipped to
  dset=dset # Variable prefix for database tables to be created
            # eg. emis or aurum or gold
)
```
</div>

## Have a cup of coffee (go for lunch)

<div style="font-size: 0.8em;">
\></br>
INFO [2026-01-16 16:54:35] emis_patient: 249961 records loaded from 1 file(s) </br>
INFO [2026-01-16 16:54:36] emis_practice: 1483 records loaded from 1 file(s)  </br>
INFO [2026-01-16 16:54:37] emis_staff: 1336287 records loaded from 1 file(s) </br>
INFO [2026-01-16 16:58:22] emis_consultation: 120456179 records loaded from 10 file(s) </br>
INFO [2026-01-16 17:17:25] emis_observation: 425006899 records loaded from 43 file(s) </br>
INFO [2026-01-16 17:17:29] emis_referral: 2826888 records loaded from 1 file(s) </br>
INFO [2026-01-16 17:17:52] emis_problem: 14501709 records loaded from 1 file(s) </br>
INFO [2026-01-16 17:27:34] emis_drugissue: 197222175 records loaded from 35 file(s) </br>
\>
</div>
- 18GB duckdb file created
- 761,601,581 records ingested in 1978 seconds (~33 minutes)
- Still think you need a database server?

## Add the Code-list

<div style="font-size: 0.8em;">
```{r load-coi_R, comment=NA, eval=F}
coi <- file.path(.projLoc,"dementia_codes.xlsx") |>
  readxl::read_excel(col_type="text") |>
  mutate(
    preferred_term=iconv(preferred_term, 
                         from="UTF-8", to="ASCII//TRANSLIT"),
    proposed_group=iconv(proposed_group, 
                         from="UTF-8", to="ASCII//TRANSLIT")
 ) 
rtrhd::load_table(dbf=dbname,dataset=coi,tab_name="coi_emis_dementia")
```
</div>

## The Database

<div style="font-size: 0.5em;">
```{r the-db_R, comment=NA,echo=F,eval=tohtml}
  rtrhd::list_db(dbname) |> plibr::display_data(buttons=F)
```

```{r the-db-md_R, comment=NA,echo=F,eval=!tohtml}
  rtrhd::list_db(dbname) |> knitr::kable(format="pipe")
```
</div>

# What Happened to T(ransform) 

## Data Quality Issues

- There will be lots of data quality issues you need to understand. (It is routinely collected healthcare
  data!)
- These are much better understood if
    - You can easily explore your data
    - You can run many look-see queries
    - You can visualise the results

# Example

## `dbplyr`

<div style="font-size: 0.8em;">
```{r ta-dah_R, comment=NA}
# connect to the database
dbcon <- DBI::dbConnect(duckdb::duckdb(),dbname,read_only=T)
# build the query
dementia_trends <- tbl(dbcon,"emis_observation") |>
  inner_join(tbl(dbcon,"coi_emis_dementia"), by = "medcodeid") |>
  group_by(patid,proposed_group) |> window_order(obsdate) |>
  summarise(
    index_date=first(obsdate),
    .groups="drop") |>
  inner_join(tbl(dbcon,"emis_patient") |>
             transmute(patid,yob,gender,regstartdate), by = "patid") |>
  mutate(index_year=year(index_date),index_age=index_year-yob) |>
# pull the data to R
  collect()
# disconnect the database
DBI::dbDisconnect(dbcon)
```
</div>

## `tibble`

<div style="font-size: 0.5em;">
```{r ta-dah_2_R, comment=NA, echo=F,eval=tohtml}
  dementia_trends |> head(10) |> mutate(patid="********") |> plibr::display_data(buttons=F)
```
</div>

```{r ta-dah_2_md_R, comment=NA, echo=F,eval=!tohtml}
  dementia_trends |> head(10) |> mutate(patid="********") |> knitr::kable(format="pipe")
```

## `ggplot2`

<div style="font-size: 0.8em;">
```{r visualise_R, comment=NA}
dementia_trends_plot <- dementia_trends %>%
  ggplot(aes(x = index_age, fill = proposed_group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Age at First Dementia Diagnosis by Type (250K patients 425M Observations)",
    x = "Age at Index Diagnosis",
    y = "Density"
  ) +
  theme_minimal()
```
</div>

## Uh oh!

```{r db_1}
tohtml
```

```{r plotit_R, comment=NA, eval=tohtml,echo=F, out.width="100%",warning=F,message=F}
print(dementia_trends_plot)
```

```{r show-plot-markdown, echo=FALSE, eval=!tohtml, results='asis'}
figpath <- "https://github.com/PietaSchofield/cprdwrangling/tree/main/README_files/figure-gfm"
figfile <- file.path(figpath,"plotit-R.png")
cat(paste0("\n\n![Dementia trends plot](",figfile,")\n\n"))
```

# This is "Lazy Evaluation" </br></br>Spoiler Alert: chain `dbplyr` then collect()

## Step 1.

<div style="font-size: 0.8em;">
```{r take2_1_R, comment=NA}
dbcon <- DBI::dbConnect(duckdb::duckdb(), dbname, read_only=T)

dementia_trends <- tbl(dbcon, "emis_observation") |>
  inner_join(tbl(dbcon, "emis_patient") |>
    select(patid, yob, gender, regstartdate, regenddate), 
    by = "patid") |>
  # Filter early in DuckDB - much more efficient
  filter(
    obsdate >= regstartdate,
    # Use today's date or end of data
    obsdate <= coalesce(regenddate, as.Date('2024-12-31'))
  ) 
```
</div>

## Step 2.

<div style="font-size: 0.8em;">
```{r take2_2_R, comment=NA}
dementia_trends <- dementia_trends |>
  inner_join(tbl(dbcon, "coi_emis_dementia"), by = "medcodeid") %>%
  group_by(patid, proposed_group) %>% 
  window_order(obsdate) %>%
  summarise(
    index_date = first(obsdate),
    .groups = "drop"
  ) %>%
  inner_join(tbl(dbcon, "emis_patient") %>% 
    transmute(patid, yob, gender, regstartdate), 
    by = "patid") 
```
</div>

## Step 3.

<div style="font-size: 0.8em;">
```{r take2_3_R, comment=NA}
dementia_trends <- dementia_trends |>
  mutate(
    index_year = year(index_date),
    index_age = index_year - yob
  ) %>%
  # Additional quality filters in DuckDB
  filter(
    index_age >= 18,
    index_age <= 110
  ) %>%
  collect()
DBI::dbDisconnect(dbcon)
```
</div>

## Replot

<div style="font-size: 0.8em;">
```{r visualise_2_R, comment=NA, out.width="100%", echo=F}
dementia_trends_plot <- dementia_trends %>%
  ggplot(aes(x = index_age, fill = proposed_group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Age at First Dementia Diagnosis by Type (250K patients 425M Observations)",
    x = "Age at Index Diagnosis",
    y = "Density"
  ) +
  theme_minimal()
```

```{r plotit_2_R, comment=NA, echo=FALSE, eval=tohtml, out.width="100%",warning=F,message=F}
print(dementia_trends_plot)
```
</div>


```{r show-plot-2-markdown, echo=FALSE, eval=!tohtml, results='asis'}
figpath <- "https://github.com/PietaSchofield/cprdwrangling/tree/main/README_files/figure-gfm"
figfile <- file.path(figpath,"plotit_2_R.png")
cat(paste0("\n\n![Dementia trends plot](",figfile,")\n\n"))
```

# Discussion

## What This Talk Was Not

- This wasn't a talk about how to clean your data
- An exhaustive tutorial on R, `tidyverse`, `dbplyr` or `duckdb`
- Dissing `data.tables` or `python` or `sql` any other tool/language

## What This Talk Was

- To show the potential of `duckdb` and `dbplyr`
- "This talk was brought to you by..."
<div style="font-size: 0.8em;">
```{r installs_R, comment=NA,eval=F}
# CRAN packages
install.packages(c("duckdb","tidyverse","dbplyr","remotes","rmarkdown"))

# My R Tools for Routine Healthcare Data (rtrhd) package 
# (and its dependencies)
remotes::install_github("PietaSchofield/rtrhd")
```
</div>
- __WARNING on Manjaro Linux duckdb takes AGES!! to install__ ... no __really__ go for dinner, watch a 
  movie contemplate, life choices long,... but on Windows and Mac I understand it is almost instant

## Take Home Message

- I loaded the data it took 33 minutes
- To render these slides using `rmarkdown::render`
    - live from the database
    - counting rows in each of 9 tables
    - queries processing 425Million records twice
    - drawing two ggplots
    - __< 15 seconds__

# Thank You (pietas@liverpool.ac.uk)

```{r build-it_R, comment=NA,eval=F,echo=F}
outfile <- file.path("/srv","http","uol","tutorials","cprdwrangling.html")
readme <- file.path(Sys.getenv("HOME"),"GitLab","cprdwrangling","README.md")
rmarkdown::render(
 input=file.path(Sys.getenv("HOME"),"GitLab","cprdwrangling","index.Rmd"),
 output_file=outfile
 )
rmarkdown::render(
 input=file.path(Sys.getenv("HOME"),"GitLab","cprdwrangling","index.Rmd"),
 output_format="github_document",
 output_file=readme
 )
system2("xdg-open",args=outfile,wait=FALSE)
system2("firefox",args=readme,wait=FALSE)
```
